---
title: "HRM Traffic Collisions"
author: "Henry Truong"
date: "11/01/2022"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = '',
                      fig.align = 'center',
                      fig.width = 7.5,
                      fig.asp = 1/1.5)
```

```{r}
library(tidyverse)
library(modelr)
library(lubridate)
library(broom)
library(infer)
library(leaflet)
library(knitr)
library(kableExtra)
library(flexdashboard)
```

```{r}
set.seed(12345)
```

```{r}
tc <- read_csv("./Traffic_Collisions.csv")
```

```{r}
tc <- tc %>% 
  rename_with(str_to_lower) %>% 
  rename(aggressive_driving = agressive_driving) %>% 
  rename_with(str_to_lower) %>% 
  mutate(date = accident_date %>% str_extract('^.{10}') %>% ymd()) %>% 
  mutate(non_fatal_injury = replace_na(
           non_fatal_injury, 
           'No'),
         fatal_injury = replace_na(
           fatal_injury, 
           'No'),
         collision_configuration = replace_na(
           collision_configuration, 
           'Not specified'),
         road_configuration = replace_na(
           road_configuration, 
           'Not specified'),
         road_location_2 = replace_na(
           road_location_2, 
           'Not specified')) %>% 
  mutate(road_location_1 = case_when(
    objectid == 2026 ~ 'Gottingen St',
    objectid == 4170 ~ 'Circassion Dr',
    objectid == 4478 ~ 'Old Guysborough Rd',
    objectid == 11536 ~ 'Osborne Ave',
    TRUE ~ road_location_1)) %>% 
  mutate(across(road_location_1:collision_configuration, 
                str_to_title)) %>% 
  mutate(year = year(date),
         month = month(date, label = TRUE),
         dow = wday(date, label = TRUE))
```

```{r plot_function}
daily_plot <- function(tc, select_year = 2021) {
  tc %>% 
    filter(year(date) == select_year) %>% 
    count(date) %>%
    ggplot(aes(x = date, y = n)) + 
    geom_line(size = 1) + 
    geom_smooth(se = F) +
    labs(x = NULL, y = 'Number of Incidents', 
         title = str_c('The Year of ', select_year)) +
    scale_x_date(date_breaks = '1 month', date_labels = '%b') +
    theme(text = element_text(size = 15))
}
```

```{r plot_function_2}
daily_timescale_barplot <- function(tc, timescale = 'year') {
  tc %>%
    mutate(time_scale = tc[[timescale]]) %>% 
    count(date, time_scale) %>% 
    group_by(time_scale) %>% 
    summarize(n_mean = mean(n)) %>% 
    ggplot(aes(x = as_factor(time_scale), y = n_mean)) +
    labs(x = NULL, y = 'Average number of daily incidents') +
    geom_col(fill = 'steelblue', width = 0.7)
}
daily_timescale_boxplot <- function(tc, timescale = 'year') {
  tc %>% 
    mutate(time_scale = tc[[timescale]]) %>% 
    count(date, time_scale) %>% 
    ggplot(aes(x = as_factor(time_scale),
               y = n)) +
    geom_boxplot() +
    stat_summary(fun = mean, color = 'darkred') +
    labs(x = NULL, y = 'Average number of daily incidents') +
    coord_flip()
}
```

```{r}
road_name <- str_split(tc$road_location_1, 
                       '^\\d*\\b', 
                       simplify = T)[, 2] %>% 
  str_trim()
```

```{r}
tc_road_name = tc %>% 
  mutate(road_name = road_name) %>% 
  filter(!str_starts(road_name, '\\d'))
```

```{r}
young_pie <- function(tc, incident = 'total') {
  if (incident == 'fatal') {
    tc_young <- tc %>% 
      filter(fatal_injury == 'Yes')
    title <- 'Regarding Fatal Incidents'
  } else if (incident == 'injury') {
    tc_young <- tc %>% 
      filter(non_fatal_injury == 'Yes' | fatal_injury == 'Yes')
    title <- 'Regarding Incidents with Injury'
  } else {
    tc_young <- tc
    title <- 'Regarding Total Incidents'
  }
  tc_young %>% 
    count(young_demographic) %>%
    mutate(percent = n/sum(n) * 100,
           ypos = cumsum(percent) - 0.5*percent) %>% 
    ggplot(aes(x = '', y = percent)) +
    geom_bar(aes(fill = fct_rev(young_demographic)), 
             stat = 'identity',
             color = 'white') +
    scale_fill_brewer(palette = 'Dark2',
                      name = NULL,
                      labels = c('People under 25 yrs old',
                                 'People at or over 25 yrs old')) +
    coord_polar(theta = 'y') +
    geom_text(aes(y = ypos, 
                  label = str_c(round(percent, 1), '%')), 
              color = "white", 
              size = 10) +
    theme_void() +
    guides(fill = guide_legend(nrow = 1)) +
    theme(legend.position = 'bottom',
          plot.title = element_text(hjust = 0.5))
}
```

```{r}
wd_we_prop_injury <- tc %>%
  group_by(date, dow) %>% 
  summarize(
    total = n(),
    injury = sum(non_fatal_injury == 'Yes' | fatal_injury == 'Yes'),
    prop_injury = injury/total) %>% 
  mutate(wd_we = case_when(
    dow == 'Sun' | dow == 'Sat' ~ 'Weekend',
    TRUE ~ 'Weekday')) %>% 
  select(wd_we, prop_injury)
```

```{r}
wd_we_boot_distribution <- wd_we_prop_injury %>% 
  specify(prop_injury ~ wd_we) %>% 
  generate(reps = 10000, type = 'bootstrap') %>% 
  calculate(stat = 'diff in means', order = c('Weekend', 'Weekday'))

wd_we_ci_percentile <- wd_we_boot_distribution %>% 
  get_ci(type = 'percentile', level = 0.95)
```

```{r}
tc_agg <- tc %>% 
  filter(aggressive_driving == 'Y',
         distracted_driving == 'N',
         impaired_driving == 'N') %>% 
  mutate(injury = case_when(
    non_fatal_injury == 'Yes' | fatal_injury == 'Yes' ~ 'Yes',
    TRUE ~ 'No'))

tc_imp <- tc %>% 
  filter(aggressive_driving == 'N',
         distracted_driving == 'N',
         impaired_driving == 'Y') %>% 
  mutate(injury = case_when(
    non_fatal_injury == 'Yes' | fatal_injury == 'Yes' ~ 'Yes',
    TRUE ~ 'No'))

agg_imp <- tibble(
  condition = c(rep('aggressive', nrow(tc_agg)),
                rep('impaired', nrow(tc_imp))),
  injury = c(tc_agg$injury, tc_imp$injury)) %>% 
  sample_n(size = nrow(tc_agg) + nrow(tc_imp))
```

```{r}
obs_diff_prop = agg_imp %>% 
  specify(injury ~ condition, success = 'Yes') %>% 
  calculate(stat = 'diff in props', 
            order = c('impaired', 'aggressive'))

agg_imp_null_distribution <- agg_imp %>% 
  specify(injury ~ condition, success = 'Yes') %>% 
  hypothesize(null = 'independence') %>% 
  generate(reps = 10000, type = 'permute') %>% 
  calculate(stat = 'diff in props', 
            order = c('impaired', 'aggressive'))

agg_imp_p_value <- agg_imp_null_distribution %>% 
  get_p_value(obs_diff_prop, direction = 'right')
```

```{r}
fri_sat_imp <- tc %>% 
  mutate(fri_sat = case_when(
    dow %in% c('Fri', 'Sat') ~ 'Friday/Saturday',
    TRUE ~ 'Other days')) %>%
  count(fri_sat, impaired_driving) 
```

```{r}
storm_date <- tribble(
  ~date, ~storm, 
  '2021-01-02', 'snow',
  '2021-01-22', 'snow',
  '2021-03-19', 'snow',
  '2020-11-03', 'snow',
  '2020-02-10', 'snow', 
  '2020-01-16', 'snow',
  '2019-09-07', 'tropical',
  '2019-09-08', 'tropical',
  '2019-09-10', 'tropical',
  '2019-09-09', 'tropical',
  '2019-09-23', 'tropical',
  '2019-09-22', 'tropical',
  '2019-11-28', 'snow',
  '2018-03-08', 'snow',
  '2018-12-07', 'snow', 
  '2018-11-16', 'snow'
)
storm_date <- storm_date %>% 
  mutate(date = ymd(date))
```

```{r}
tc_model_data <- left_join(tc, storm_date, by = 'date') %>% 
  mutate(storm = replace_na(storm, 'normal')) %>% 
  group_by(date, dow, month, storm) %>% 
  summarize(n = n())
```

```{r}
model1 <- lm(n ~ dow, data = tc_model_data)
model2 <- lm(n ~ dow + storm, data = tc_model_data)
model3 <- lm(n ~ dow + storm + month, data = tc_model_data)
list_models <- list(dow = model1, 
                   dow_with_storm = model2,
                   dow_with_storm_month = model3)
```


Time
===================================== 

Column {data-width=700}
-----------------------------------------------------------------------

```{r}
selectInput("year", label = "Select a year",
            choices = c(2018, 2019, 2020, 2021), selected = 2021)
```

### Daily incident number in the selected year

```{r}
renderPlot({
  daily_plot(tc, input$year)
})
```

Column {data-width=300}
-----------------------------------------------------------------------

### Average daily incident number in 2018-2021

#### 

```{r, fig.width=4, fig.height=2}
daily_timescale_barplot(tc, 'year')
```

#### 

```{r, fig.width=4}
daily_timescale_boxplot(tc, 'year')
```

Time Scale
===================================== 

Column {data-width=330}
-----------------------------------------------------------------------

```{r}
radioButtons('timescale', 'Select a timescale',
             choiceNames = c('Month', 'Day of week'),
             choiceValues = c('month', 'dow'),
             inline = TRUE)
```


### Average daily incident number among the selected timescale

```{r, fig.height=10, fig.width=5}
renderPlot({
  time_scale_plot <- daily_timescale_barplot(tc, input$timescale) + 
    coord_flip() +
    theme(text = element_text(size = 15))
  if (input$timescale == 'month') {
    time_scale_plot + 
      geom_col(fill = c(rep('steelblue', 3),
                        'darkolivegreen',
                        rep('steelblue', 7),
                        'darkred'))
  } else {
    time_scale_plot +
      geom_col(fill = c('darkolivegreen',
                        rep('steelblue', 4),
                        'darkred',
                        'steelblue'))
  }
   
})
```

Column {data-width=670}
-----------------------------------------------------------------------

### Distribution of daily incident number {data-height=200}

```{r, fig.asp=1/2}
tc %>% 
  count(date) %>% 
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 3, boundary = 0, 
                 color = 'white', fill = 'steelblue') +
  scale_x_continuous(breaks = seq(0, 90, by = 9)) +
  labs(x = "Daily incident number",
       y = "Count") +
  theme(text = element_text(size = 9))
```


### Summary table of daily incident number {data-height=27}
```{r}
renderTable({
  summary_table <- tc %>% 
    count(date) %>% 
    .$n %>% 
    summary() %>% 
    as.list() %>% 
    data.frame()
  colnames(summary_table) <- c('Minimum', '1st Quantile', 
                               'Median', 'Mean',
                               '3rd Quantile', 'Maximum')
  summary_table
})
```

Geography 
===================================== 

Column {data-width=670}
-----------------------------------------------------------------------

### Distribution of incidents across HRM

```{r}
tc %>% 
  select(lat = wgs84_lat_coord, 
         lng = wgs84_lon_coord) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(clusterOptions = markerClusterOptions())
```

Row {data-width=330 .tabset}
-----------------------------------------------------------------------

```{r}
wellPanel(fluidRow(column(
  7,
  sliderInput('n', 'Top N Roads', 
            min = 3, max = 10, 
            step = 1, value = 10,)
)))
```

### Total incidents

```{r, fig.asp=2}
renderPlot({
  tc_road_name %>% 
  count(road_name) %>% 
  arrange(desc(n)) %>% 
  slice(1:input$n) %>% 
  ggplot(aes(x = fct_reorder(road_name, n), y = n)) +
  geom_col(fill = 'steelblue', width = 0.9*input$n/10) +
  labs(x = NULL, y = "Number of Incidents") +
  coord_flip() +
  theme(text = element_text(size = 15))
})
```

### Incidents with injury

```{r}
renderPlot({
  tc_road_name %>% 
  filter(non_fatal_injury == 'Yes' | fatal_injury == 'Yes') %>% 
  count(road_name) %>% 
  arrange(desc(n)) %>% 
  slice(1:input$n) %>% 
  ggplot(aes(x = fct_reorder(road_name, n), y = n)) +
  geom_col(fill = 'steelblue', width = 0.9*input$n/10) +
  labs(x = NULL, y = "Number of Incidents with Injury") +
  coord_flip() +
  theme(text = element_text(size = 15))
})
```

### Fatal incidents

```{r}
renderPlot({
  tc_road_name %>% 
  filter(fatal_injury == 'Yes') %>% 
  count(road_name) %>% 
  arrange(desc(n)) %>% 
  slice(1:input$n) %>% 
  ggplot(aes(x = fct_reorder(road_name, n), y = n)) +
  geom_col(fill = 'steelblue', width = 0.9*input$n/10) +
  labs(x = NULL, y = "Number of Fatal Incidents") +
  coord_flip() +
  theme(text = element_text(size = 15))
})
```

Collision and Driver Characteristics
===================================== 

Column {data-width=600, .tabset}
-------------------------------------

### Total incidents

```{r}
tc %>% 
  filter(collision_configuration != 'Not Specified') %>% 
  count(collision_configuration) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = fct_reorder(collision_configuration, n), y = n)) +
  geom_col(fill = 'steelblue') +
  labs(x = NULL, y = "Number of Incidents") +
  coord_flip() +
  theme(text = element_text(size = 12))
```

### Incidents with injury

```{r}
tc %>% 
  filter(collision_configuration != 'Not Specified') %>% 
  filter(non_fatal_injury == 'Yes' | fatal_injury == 'Yes') %>% 
  count(collision_configuration) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = fct_reorder(collision_configuration, n), y = n)) +
  geom_col(fill = 'steelblue') +
  labs(x = NULL, y = "Number of Incidents with Injury") +
  coord_flip() +
  theme(text = element_text(size = 12))
```

### Fatal incidents

```{r}
tc %>% 
  filter(collision_configuration != 'Not Specified') %>% 
  filter(fatal_injury == 'Yes') %>% 
  count(collision_configuration) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = fct_reorder(collision_configuration, n), y = n)) +
  geom_col(fill = 'steelblue') +
  labs(x = NULL, y = "Number of Fatal Incidents") +
  coord_flip() +
  theme(text = element_text(size = 12))
```

Column {data-width=400 .tabset}
-------------------------------------

### Total incidents

```{r, fig.asp=1}
young_pie(tc, 'total') +
  theme(text = element_text(size = 20))
```

### Incidents with injury

```{r, fig.asp=1}
young_pie(tc, 'injury') +
  theme(text = element_text(size = 20))
```

### Fatal incidents

```{r, fig.asp=1}
young_pie(tc, 'fatal') +
  theme(text = element_text(size = 20))
```

Statistical Inference 
===================================== 

```{r}
selectInput(
    'con_lev', 'Select the level of confidence',
    c(0.90, 0.95, 0.99)
    )
```

Column {data-width=400}
-------------------------------------

### Data Visualization {data-height=500}
```{r, fig.asp=1/1.5}
wd_we_prop_injury %>% 
  ggplot(aes(x = wd_we, y = prop_injury)) +
  geom_boxplot() +
  stat_summary(fun = mean, color = 'darkred', size = 0.5) +
  labs(x = NULL, y = "Proportion of Incidents with Injury") +
  coord_flip() + 
  theme(text = element_text(size = 15))
```

### Confident interval {data-height=150}
```{r}
renderTable(
  wd_we_ci_percentile <- wd_we_boot_distribution %>% 
    get_ci(type = 'percentile', level = as.numeric(input$con_lev))
)
```

Column {data-width=600, data-height=600}
-------------------------------------

### Bootstrap Distribution
```{r}
renderPlot({
  wd_we_ci_percentile <- wd_we_boot_distribution %>% 
    get_ci(type = 'percentile', level = as.numeric(input$con_lev))
  wd_we_boot_distribution %>%
  visualize() +
  shade_ci(wd_we_ci_percentile) +
  labs(x = 'Diffence in Mean Proportions between\n Weekday and Weekend',
       y = 'Count',
       title = str_c('The ', input$con_lev, ' Confident Interval Contains 0')) + 
  theme(text = element_text(size = 15))
})
```


Predictive Model
===================================== 

Column
-------------------------------------

### Model 1: Day of week

```{r, fig.asp= 1/3}
tc_model_data %>% 
  gather_residuals(dow = model1, 
                   dow_with_storm = model2, 
                   dow_with_storm_month = model3) %>% 
  filter(model == 'dow') %>% 
  ggplot(aes(date, resid)) +
  geom_line() + 
  geom_smooth(se = F) +
  labs(x = "Date", y = "Residual") +
  geom_hline(yintercept = 0, size = 1, colour = "white")
```

### Model 2: Day of week and severe weather

```{r, fig.asp=1/3}
tc_model_data %>% 
  gather_residuals(dow = model1, 
                   dow_with_storm = model2, 
                   dow_with_storm_month = model3) %>% 
  filter(model == 'dow_with_storm') %>% 
  ggplot(aes(date, resid)) +
  geom_line() + 
  geom_smooth(se = F) +
  labs(x = "Date", y = "Residual") +
  geom_hline(yintercept = 0, size = 1, colour = "white")
```

Column
-------------------------------------

### Model 3: Day of week, severe weather and month {data-height=500}

```{r, fig.asp=1/3, fig.width=7.5}
tc_model_data %>% 
  gather_residuals(dow = model1, 
                   dow_with_storm = model2, 
                   dow_with_storm_month = model3) %>% 
  filter(model == 'dow_with_storm_month') %>% 
  ggplot(aes(date, resid)) +
  geom_line() + 
  geom_smooth(se = F) +
  labs(x = "Date", y = "Residual") +
  geom_hline(yintercept = 0, size = 1, colour = "white")
```

### Model and R-squared summary {data-height=500}

####
```{r, echo=T}
model1 <- lm(n ~ dow, data = tc_model_data)
model2 <- lm(n ~ dow + storm, data = tc_model_data)
model3 <- lm(n ~ dow + storm + month, data = tc_model_data)
```

####
```{r}
renderTable(
  list_models %>% 
  map_dbl(~ summary(.)[['r.squared']]) %>% 
  enframe(name = 'model', value = 'r_squared')
)
```





